---
title: 'Part 1: Background and Basics'
author: "Michael Kane, Yale University"
output: 
  ioslides_presentation:
    theme: flatly
---

# Welcome to Bootstrap Learning Python with R

## What are the goals of this class?

Learn how to translate your R programming skills to other langauges, Python in particular.

Separate programming language _syntax_ from programming _constructs_.

Appreciate the linguistic differences between R and Python and think about the implications for expressing computations and execution efficiency.

Be able to evaluate objective strengths and weaknesses of a programming language.

## How will this class be taught?

I'm going to start with a core-programming constructs in R. I'll describe it, show how it works, and talk about some of it's distinguishing characteristics

I'll show the equivalent/analogous construct in Python. I'll describe it, show how it works, and we'll discuss some of the differences

You will follow along.

You execute the code, modify it, maybe (probably) break it, and ask questions.

## Who am I and why should you listen to me?

My name is Michael Kane, I'm an Assistant Professor of Biostatistics at Yale University and I work on methods in optimization-based machine learning, numerical and statistical computing.

I have a degree in Computer Engineering, master's degrees in Electrical Engineering and Statistics, and a Ph.D. in Statistics.

I've been a software architect and developer for a couple of different start-ups and other businesses. I built software in ~15 different programming languages.

## Who am I and why should you listen to me? (cont'd)

I won the 2010 ASA Chamber's Statistical Software Award for the Bigmemory Project; a package that allows R to accomodate and compute with large matrices.

I have ~7 package on [CRAN](https://cran.r-project.org/) and have developed packages (in Python and R) featured in [DARPA's Open Software Catalog](https://www.darpa.mil/opencatalog?PP=49).

I'm a founding member and of a start-up specializing in clinical trial data analyses where I architect and develop key (mostly cloud-based) analysis infrastructure.

# Becoming a Polyglot Programmer

## Why learn Python 

"Use the best tool for the job" is great advice if you know how to use all of the tools.

Python is a great general-purpose programming language.
- It is [popular](https://www.tiobe.com/tiobe-index/).
- Simple syntax
- Has a lot of facilities in non-data-science applications.

You learn to separate syntax from construct.

It will make you a better programmer.

## Is Python Better than R?

If you are looking for a [head-to-head comparison by a computer scientist, Norm Matloff](https://github.com/matloff/R-vs.-Python-for-Data-Science) does a good job.

I like avoiding the questions about language + community in favor of characterizing the language.

Python
- Object-oriented focused (manages complexity in large sofware systems - Tensorflow)
- Readable

R
- A pragmatic version of Lisp
- More dynamic (better meta-programming capabilities - DSD's)

Both of them are [Turing complete](https://en.wikipedia.org/wiki/Turing_completeness#Non-mathematical_usage), most comparisons are about areas of emphasis of the community.

# Materials and Setup

## Python Virtual Environments

"Python, like most other modern programming languages, has its own unique way of downloading, storing, and resolving packages (or modules). While this has its advantages, there were some _interesting_ decisions made about package storage, resolution [and compatibility], which has lead to some problems—particularly with how and where packages are stored."" - [realpython](https://realpython.com/python-virtual-environments-a-primer/) (emphasis added)

## setup-p4rp.sh {.smaller}

```{bash, eval = FALSE}
# Make sure python 3 and pip3 are installed on you machine and accessible from the
# command line.

# Install virtualenv.
pip3 install virtualenv

# Create a "python for R programmer's" virtual environment.
virtualenv -p python3 p4rp

# Activate the virtual environment. Note: this works for Mac and Linux.
source p4rp/bin/activate

# The following *may* work on Windows.
# p4rp\bin\activate.bat

# Install the needed packges.
pip3 install pandas numpy
```

## Running Python and R from RStudio {.smaller}

Tell RStudio/RMarkdown to use the `reticulate` package and load our newly created virtual environment.

```{r comment="", echo = FALSE}
cat("```{r}\n# R\nlibrary(reticulate)\nuse_virtualenv(\"./p4rp\")\n```")
```

```{r echo = FALSE}
library(reticulate)
use_virtualenv("./p4rp")
```

Now check that it works:

```{r comment="", echo = FALSE}
cat("```{python}\n# Python\nprint(\"Hello world from Python!\")\n```")
```

```{python echo = FALSE}
print("Hello world from Python!")
```

# The Basics

## Style (according to [PEP-8](https://www.python.org/dev/peps/pep-0008))

Python doesn't use { } to enclose code blocks it uses indentation. _[Four spaces](https://www.python.org/dev/peps/pep-0008/#indentation)_ is recommended.

[Line length should be limited to 72 characters](https://www.python.org/dev/peps/pep-0008/#maximum-line-length).

[Functions and variables should use lowercase with words separated by underscores as necessary to improve readability](https://www.python.org/dev/peps/pep-0008/#method-names-and-instance-variables).

Variables, types, and functions cannot contain "." This is an infix operator in Python and it's used to call methods (functions) and member variables.

## Getting help

Python's help function is `help()`. R's help function is the prefix operator `?`.

```{r eval = FALSE}
# R
# Get help on the ? function
?`?`
```

```{python eval = FALSE}
# Python
# Get help on the help function
help(help)
```

## Variables and Assignment

In Python assignments are performed using `=`. There is no `<-`

```{r}
# R
a <- 3
print(a + 2)
```

```{python}
# Python
a = 3
print(a + 2)
```

## Basic types {.smaller}

```{r}
# R
# Note: factor and raw not included
logical <- TRUE
numeric <- 2.72
integer <- 42L
complex <- 3 + 2i
character <- "string"
cat(logical, numeric, integer, complex, character)
```

```{python}
# Python
logical = True
numeric = 2.72
integer = 42
complex = 3 + 2j # Note: j... like an engineer.
character = "string"
print(logical, numeric, integer, complex, character)
```

## What about NA, NaN, Null, -Inf, and Inf? {.smaller}

Python has `None`

- `NA` and others need to be imported from packages. 
- It is rougly equivalent to `Null`.
- We'll get to these later.

"Use `is` when you want to check against an object's identity (e.g. checking to see if var is None). Use `==` when you want to check equality (e.g. Is var equal to 3?)." - [Stack Overflow User](https://stackoverflow.com/questions/14247373/python-none-comparison-should-i-use-is-or#14247383)

```{python}
#Python
a = None
print(a is None)
```


# Control Structures and Functions

## if Statements {.smaller}

```{r}
# R
confused <- FALSE
if (!confused) {
  cat("I'm getting it.")
} else {
  cat("Huh?")
}
```

```{python}
# Python
confused = False
if not confused:
    print("I'm getting it.")
else:
    print("Huh?")
```

## `if-elif-else` {.smaller}

```{python}
# Python
confusion_level = 0 # assume confusion is 0 to 10 (very confused)
if 0 <= confusion_level < 3:
    print("I'm getting it.")
elif 3 <= confusion_level < 6:
    print("I'm kind of getting it.")
elif 6 < confusion_level <= 10:
    print("I'm not really getting it.")
else:
    print("Where am I?")
```

## `for` Loops {.smaller}

```{R}
# R
for (i in c(1, 2, 3)) {
  print(i)
}
```

```{python}
# Python
for i in (1, 2, 3):
    print(i)
```

## Another `for` Loop {.smaller}

```{R}
# R
for (i in seq_len(3)) {
  print(i)
}
```

```{python}
# Python
for i in range(3):
    print(i+1)
```

## `while` Loops {.smaller}

```{R}
# R
i <- 1
while (i < 4) {
  print(i)
  i <- i + 1
}
```

```{python}
# Python
i = 1
while i < 4:
    print(i)
    i += 1
```

## Functions

```{r}
# R
add_one <- function(v) {
  v + 1
}

cat(add_one(3))
```

```{python}
# Python
def add_one(v):
    return v + 1
    
print(add_one(3))
```

## Don't forget to return values you want

```{python}
# Python
def add_one(v):
    ret = v + 1
    print(ret)
    ret
    
print(add_one(3))
```

# Lists

## R Lists {.smaller}

```{r}
# R
v <- list(1, 2)
print(v)
```

```{r}
# R
v <- c(v, list(3))
print(v)
```

## R Lists Continued {.smaller}

```{r}
# R
print(v[-1])
```

## Python Lists {.smaller}

```{python}
# Python
v = [1, 2]
print(v)
```

```{python}
# Python
v = [1, 2]
print(v + [3])
```

## Python Lists Continued {.smaller}

```{python}
# Python
v = [1, 2, 3]
print(v[-1])
```

## List Indexing {.smaller}

R indexes start at 1. Python starts at 0.

```{r}
# R
v <- c('a', 'b', 'c')[3]
cat(v)
```

```{python}
# Python
v = ['a', 'b', 'c']
print(v[2])
```

## List Range Indexing {.smaller}

R indexes from the first argument _through_ the second. Python indexes from the first _up to_ the last (non-inclusive).

```{r}
# R
v <- c('a', 'b', 'c')[1:2]
cat(v)
```

```{python}
# Python
v = ['a', 'b', 'c']
print(v[0:2])
print(v[:2])
```

## Lists can contain values with different types {.smaller}

```{python}
# Python
list_vals = [1, '2', None]

def add_one(list_vals):
    for i in range(len(list_vals)):
        if list_vals[i] is None:
            print("Value at index " + str(i) + " is None. Not adding 1.")
        elif isinstance(list_vals[i], str):
            list_vals[i] = str(int(list_vals[i]) + 1)
        elif isinstance(list_vals[i], int):
            list_vals[i] += 1
        else:
            print("Unsupported type at ", i, ". Not doing anything")
    return list_vals

print(list_vals)
print(add_one(list_vals))
print(list_vals)
```

## Values are passed _by reference_ {.smaller}

```{python}
# Python
list_vals = [1, '2', None]

def add_one(list_vals):
    ret_vals = list_vals.copy()
    for i in range(len(ret_vals)):
        if ret_vals[i] is None:
            print("Value at index " + str(i) + " is None. Not adding 1.")
        elif isinstance(ret_vals[i], str):
            ret_vals[i] = str(int(ret_vals[i]) + 1)
        elif isinstance(ret_vals[i], int):
            ret_vals[i] += 1
        else:
            print("Unsupported type at ", i, ". Not doing anything")
    return ret_vals

lv_one_added = add_one(list_vals)
print(lv_one_added)
print(list_vals)
```

## `=` Doesn't copy lists

```{python}
# Python
a = [3]
b = a
b[0] = 4
print(a)
```

## It does copy singletons

```{python}
# Python
a = 3
b = a
b = 4
print(a)
```

## Lists are not like R vectors {.smaller}

```{r}
# R
a <- 1:3
print(a + 1)
```

```{python error = TRUE}
# Python
a = list(range(1, 4))
print(a)
a + 1
```
## List Comprehensions {.smaller}

```{r}
# R
l <- as.list(1:3)
print(lapply(l, function(x) x + 1))
```

```{python}
# Python
l = list(range(1, 4))
print( [x + 1 for x in l] )
```

## A note on Tuples {.smaller}

In Python a tuple is an _immutable_ list. 

```{python error = TRUE}
t1 = 1, 2, 3   # Create a tuple like this...
t2 = (1, 2, 3) # or like this

t1[1] = 2 # Throws an error

first_equal, second_equal, third_equal = [ t1[i] == t2[i] for i in range(len(t1)) ]
print(first_equal, second_equal, third_equal)
```

# Dictionaries 

## Dictionaries in Python are named lists in R) {.smaller}

```{r}
# R
l <- list(a = c(1:3), b = c("c", "d", "e", "f"))
l$b[2:4]
```

```{python}
# Python
l = {'a' : list(range(1, 4)),
     'b' : ["c", "d", "e", "f"]
     }
     
print(l['b'][1:4])
```

## Dictionaries are composed of keys and values {.smaller}

```{python}
# Python
l = {'a' : list(range(1, 4)),
     'b' : ["c", "d", "e", "f"]
    }

print(l.keys())
print(l.values())
```
