---
title: 'Part 3: Building Software Systems using Object-Oriented Programming'
author: "Michael Kane, Yale University"
output: 
  ioslides_presentation:
    theme: flatly
    smaller: true
---

```{r echo = FALSE}
library(reticulate)
use_virtualenv("./p4rp")
```

# Overview of Part 3 {.smaller}

## Highlights from Part 2

### Another look at environments and packages

### Numpy for R's vector/matrix/array operations

### How to use Python Objects

### Some Visualization

## Feedback Questions

### Is Python _that_ different from R?

### What do you like most about Python so far?

### Which Python constructs would you like to see in R?

## Topics for this part

### Function with State: R Mutable Closures and Python Generators

### R and Python's dispatch

### Python Classes and Objects

### Building an OLS Learner

# Function with State: R Mutable Closures and Python Generators

## In R Function with State are called _Mutable Closures_

```{R}
#R

increment_generator <- function(start = 1) {
  val <- start - 1
  function() {
    val <<- val + 1
    val
  }
}

inc <- increment_generator()
inc()
inc()
```

## How does this work?

```{R}
# R

ls(environment(inc))
environment(inc)$val
b <- inc()
environment(inc)$val
```

## So what?

### This is very useful for iterating over things

- in loops it means we don't have to have all of the values we are iterating over up front - we can iterate over things that are bigger than memory
- we can easily change the size of the things we are iterating over - useful in parallel computing where number of processors varies from machine to machine.

## This is the basis for iterators in R

```{r}
library(iterators)

icount
```

## This is the basis for iterators in R

```{r, error = TRUE}
it <- icount(2)
nextElem(it)
nextElem(it)
nextElem(it)
nextElem(it)
```

## Which can be used with the `%foreach%` packge

```{r}
library(foreach)
registerDoSEQ()

foreach(it = icount(10), .combine = c) %dopar% {
  it + 1
}
```

## ... and is an excellent package for parallelization

```{r}
library(doParallel)
registerDoParallel()

# Same code. Runs in parallel.
foreach(it = icount(10), .combine = c) %dopar% {
  it + 1
}
```

## R's iterators ~ Python's generators

```{python, error = TRUE}
# Python

def icount(n = None):
    if n is None or not isinstance(n, int):
        raise ValueError("n must be an integer.")
    i = 0
    while n is None or i < n:
        yield i
        i += 1
```

## R's iterators ~ Python's generators

```{python, error = TRUE}
# Python

it = icount(2)
next(it)
next(it)
next(it)
```

## Generators and Loops

```{python, error = TRUE}
# Python

[x + 1 for x in icount(10)]

for x in icount(10):
  print(x)
```

## Parallelizing Loops

```{python}
# Python

from multiprocessing import Pool

def add_one(x):
    return x + 1

p = Pool(processes = 4)
p.map(add_one, icount(10))
```

# R and Python's Dispatches

## What is Dispatch?

### The choice of which version of a method to call.

- If this choice is based on a single object, it is _single dispatch_
- For multiple object - _multiple dispatch_.

### A function does not dispatch.

### In R: 
- Single dispatch is supported by S3, R5 (RC) (also implemented in packges R6 and R.oo).
- Multiple dispatch is supported by S4.

### In Python
- Single dispatch is handled with objects.
- Multiple dispatch is available in a few packages.

## A quick review of R's S3 dispatch

```{r, error = TRUE, eval = FALSE}
# R

add_one <- function(x) {
  UseMethod("add_one", x)
}

add_one.default <- function(x) {
  stop(paste("Don't know how to add_one to object of type", class(x)))
}

add_one.numeric <- function(x) {
  print("Dispatched to `add_one.numeric`.")
  x
}

cat("Calling add_one on a.")
add_one("a")

cat("Calling foo on the number 1.")
add_one(1)
```

## A quick review of R's S3 dispatch (cont'd)

```{r, error = TRUE, eval = TRUE, echo = FALSE}
# R

add_one <- function(x) {
  UseMethod("add_one", x)
}

add_one.default <- function(x) {
  stop(paste("Don't know how to add_one to object of type", class(x)))
}

add_one.numeric <- function(x) {
  cat("Dispatched to `add_one.numeric`.")
  x + 1
}

cat("Calling add_one on \"a\".")
add_one("a")

cat("Calling add_one on the number 1.")
add_one(1:10)
```

## S3 in Practice

```{r, error = TRUE, eval = TRUE}
print_methods <- methods(print)
print(head(print_methods, 20))
```

## A Python Equivalent

### We already know how to do this with either lists or numpy arrays.

### The interesting part is not adding one. It is building objects that perform an operation with a common name for different types.

## We are going to start by building a _class_ which describes types of objects.

## The `AddOneToNumericList` Class

```{python}
# Python

class AddOneToNumericList:
    def __init__(self, lst):
        if any( [not isinstance(x, (float, int)) for x in lst] ):
            raise TypeError("All list elements must be int or float")
        self.lst = lst
    def add_one(self):
        self.lst = [x + 1 for x in self.lst]
    def get_lst(self):
        return(self.lst)
```

## Creating an Instance

```{python}
# Python

my_new_object = AddOneToNumericList(list(range(1, 11)))

print(my_new_object.get_lst())

print(my_new_object.lst)

my_new_object.add_one()

print(my_new_object.lst)

```

## Limiting access to attributes

```{python}
# Python

class AddOneToNumericList:
    def __init__(self, lst):
        if any( [not isinstance(x, (float, int)) for x in lst] ):
            raise TypeError("All list elements must be int or float")
        self.__lst = lst
    def add_one(self):
        self.__lst = [x + 1 for x in self.__lst]
    def get_lst(self):
        return(self.__lst.copy())
```

## Limiting access to attributes (cont'd)

```{python, error = TRUE}
# Python

my_new_object = AddOneToNumericList(list(range(1, 11)))

print(my_new_object.get_lst())

print(my_new_object.__lst)
```

## Let's abstract this a little bit

### The class works for numeric (int and float) types.

### We've already implemented adding one other other types (strings).

## Let's Create an _Abstract Class_ `AddOneToList`

```{python, eval = FALSE, error = TRUE}
# Python

from abc import ABC, abstractmethod 

class AddOneToList(ABC):
    @abstractmethod
    def __init__(self, lst):
        self._lst = lst

    @abstractmethod
    def add_one(self):
        pass
        
    @abstractmethod
    def get_lst(self):
        pass

ao = AddOneToList(list(range(1, 11)))       
```

## Let's Create an _Abstract Class_ `AddOneToList` (cont'd)

```{python, echo = FALSE, error = TRUE}
# Python

from abc import ABC, abstractmethod 

class AddOneToList(ABC):
    @abstractmethod
    def __init__(self, lst):
        self._lst = lst

    @abstractmethod
    def add_one(self):
        pass
    @abstractmethod
    def get_lst(self):
        pass

ao = AddOneToList(list(range(1, 11)))       
```

## `@abstractmethod` is a _Decorator_

```{python}
def check_second_arg_not_zero(func):
    def inner(a1, a2):
        if a2 == 0:
            print("Can't divide by zero!")
            return(None)
        return func(a1, a2)
    return inner

@check_second_arg_not_zero
def divide(num, denom):
    return num / denom
    
divide(22, 7)
divide(22, 0)
```

## Now Let's Create `AddOneToNumericList`

```{python, error = TRUE}
# Python

class AddOneToNumericList(AddOneToList):

    def add_one(self):
        self._lst = [x + 1 for x in self._lst]

    def get_lst(self):
        return(self._lst.copy())
 
ao = AddOneToNumericList(list(range(1, 11)))       
```

## Now Create a Concrete Classes

```{python, error = FALSE}
# Python

class AddOneToIntList(AddOneToNumericList):

    def __init__(self, lst):
        if any( [not isinstance(x, int) for x in lst] ):
            raise TypeError("All list elements must be int!")
        super().__init__(lst)

aoi = AddOneToIntList(list(range(1, 11)))
aoi.add_one()
print(aoi.get_lst())
```

## Now Create a Concrete Classes (cont'd)

```{python, error = TRUE}
# Python

AddOneToIntList([float(x) for x in range(1, 11)])
```

## What if I want to support a bunch of different types and let the constructor figure out which one to make?

### You want a _factory_

### You provide a list, it figures out if it can create an instance of a class derived from AddOneToList and gives it to you.

### You don't even need to specify the concrete types before hand.

### We'll need to revise `AddOneToList`.

## `AddOneToList` Redux

```{python, error = FALSE}
# Python

def get_lst_type(lst):
    if len(lst) == 0:
        raise AssertionError("List length is zero.")
    lst_types = [type(lst[0]) == type(x) for x in lst]
    return type(lst[0])
```

## `AddOneToList` Redux (cont'd)

```{python, error = FALSE}
# Python

class AddOneToList(ABC):

    def factory(lst):
        if get_lst_type(lst) is int:
            return AddOneToIntList(lst)
        else:
            raise TypeError("Unsupported list type.")
    
    factory = staticmethod(factory) 

    @abstractmethod
    def __init__(self, lst):
        self._lst = lst

    @abstractmethod
    def add_one(self):
        pass
    @abstractmethod
    def get_lst(self):
        pass
```


## `AddOneToList` Redux (cont'd)

```{python, error = FALSE}
lsts = [AddOneToList.factory(list(range(x))) for x in range(1, 3)]
print(lsts)
print(lsts[1].get_lst())
[x.add_one() for x in lsts]
print(lsts[1].get_lst())
```

## What's the point again?

### Factories create the "right" class with a little bit of information.

- they can be static, as in our example
- or they can be dynamic, allowing users to register new concrete classes

### Methods have the same interface and do different things based on class.

### Examples

- data importing, a class might know how to get data from various sources
- graphics, a class might know how to create a visualization
- model fitting, a class might know how to fit a data set


