---
title: 'Part 1: Background and Basics'
author: "Michael Kane<br><span style = 'font-size: 80%;'>School of Public Health, Biostatistics Department<br>Yale University<br><br>"
date: '`r icon::fa("github")` [kaneplusplus](https://github.com/kaneplusplus)<br> `r icon::fa("twitter")` [kaneplusplus](https://twitter.com/kaneplusplus)'
output:
  xaringan::moon_reader:
    css: [duke-blue, metropolis, tamu-fonts]
    titleSlideClass: [left, middle]
---

```{r echo = FALSE}
library(reticulate)
xaringanExtra::use_panelset()
use_condaenv("python-for-r-developers")
```

# <br> Welcome to Bootstrap Learning Python with R

<br>

## What are the goals of this class?

<br>

--

Learn how to translate your R programming skills to other langauges, Python in particular.

--

Separate programming language _syntax_ from programming _constructs_.

--

Appreciate the linguistic differences between R and Python and think about the implications for expressing computations and execution efficiency.

--

Be able to evaluate objective strengths and weaknesses of a programming language.

---

# <br> Welcome to Bootstrap Learning Python with R

<br>

## How will this class be taught?

<br>

--

I'm going to start with a core-programming constructs in R. I'll describe it, show how it works, and talk about some of it's distinguishing characteristics

--

I'll show the equivalent/analogous construct in Python. I'll describe it, show how it works, and we'll discuss some of the differences

--

You will follow along.

--

You execute the code, modify it, maybe (probably) break it, and ask questions.

---

# <br> Welcome to Bootstrap Learning Python with R

<br>

## <br> Who am I and why should you listen to me?

<br>

My name is Michael Kane, I'm an Assistant Professor of Biostatistics at Yale University and I work on methods in optimization-based machine learning, numerical and statistical computing.

--

I have a degree in Computer Engineering, master's degrees in Electrical Engineering and Statistics, and a Ph.D. in Statistics.

--

I've been a software architect and developer for a couple of different start-ups and other businesses. I built software in ~15 different programming languages.

---

# <br> What are the parts of this class?

<br>

## Background and Basics

<br>

## Digging Deeper into Python Programming Constructs

<br>

## Building Software Systems using Object-Oriented Programming

<br>

## Working with Data
---

# <br> Becoming a Polyglot Programmer

<br>

## Why learn Python 

<br>

--

"Use the best tool for the job" is great advice if you know how to use all of the tools.

--

Python is a great general-purpose programming language.

- It is [popular](https://www.tiobe.com/tiobe-index/)
- Simple syntax
- A lot of facilities in non-data-science applications

--

You learn to separate syntax from construct.

--

It will make you a better programmer.

---

# <br> Becoming a Polyglot Programmer

<br>

## Is Python Better than R?

<br>

--

If you are looking for a [head-to-head comparison by a computer scientist, Norm Matloff](https://github.com/matloff/R-vs.-Python-for-Data-Science) does a good job.


--

<br> 

I like avoiding the questions about language + community in favor of characterizing the language.

---

# <br> Becoming a Polyglot Programmer

<br>

## Is Python Better than R?

<br>

--

Python
- Object-oriented focused (manages complexity in large software systems - Keras + Tensorflow)
- Readable

--

R
- A pragmatic version of Lisp
- More dynamic (better meta-programming capabilities - DSD's)

--

Both of them are [Turing complete](https://en.wikipedia.org/wiki/Turing_completeness#Non-mathematical_usage), most comparisons are about areas of emphasis of the community.

---

# <br> Materials and Setup

<br>

## Python Virtual Environments

<br>

--

"Python, like most other modern programming languages, has its own unique way of downloading, storing, and resolving packages (or modules). While this has its advantages, there were some _interesting_ decisions made about package storage, resolution [and compatibility], which has lead to some problems—particularly with how and where packages are stored." 

-[realpython](https://realpython.com/python-virtual-environments-a-primer/) (emphasis added)


---

# <br> Materials and Setup

<br>

## What are my options for running Python?

<br>

1. The command line

--

2. RStudio

--

3. RMarkdown

--

4. IPython notebooks

---

# <br> Materials and Setup

<br>

## How do we install Python and manage packages?

--

<br>

## We use R

<br>

1. Install the `reticulate`.

--

2. Use `reticulate::install_miniconda()` to create a conda installation.

--

3. Create a conda environment with `conda_create()`

--

4. Start the conda environment.

--


See [python-and-conda-install.r](python-and-conda-install.r) for details.

---

# <br> Start the conda environment, run from R Markdown

<br>

```{r echo = TRUE, eval = FALSE}
library(reticulate)
use_condaenv("python-for-r-developers", required = TRUE)
```

--

Now check that it works:

```{r comment="", echo = FALSE}
cat("```{python}\n# Python\nprint(\"Hello world from Python!\")\n```")
```

```{python echo = TRUE}
print("Hello world from Python!")
```

---

# <br> The Basics

<br>

## Style (according to [PEP-8](https://www.python.org/dev/peps/pep-0008))

--

<br>

Python doesn't use { } to enclose code blocks it uses indentation. _[Four spaces](https://www.python.org/dev/peps/pep-0008/#indentation)_ is recommended.

--

[Line length should be limited to 72 characters](https://www.python.org/dev/peps/pep-0008/#maximum-line-length).

--

[Functions and variables should use lowercase with words separated by underscores as necessary to improve readability](https://www.python.org/dev/peps/pep-0008/#method-names-and-instance-variables).

--

Variables, types, and functions cannot contain "." This is an infix operator in Python and it's used to call methods (functions) and member attributes.

---

# <br> The Basics

<br>

## Getting help

<br>

R's help function is the prefix operator `?`.

```{r eval = FALSE}
# R
# Get help on the ? function
?`?`
```

--

<br>

Python's help function is `help()`. 

```{python eval = FALSE}
# Python
# Get help on the help function
help(help)
```

---

# <br> The Basics

<br>

## Variables and Assignment

--

```{r}
# R
a <- 3
print(a + 2)
```

--

In Python assignments are performed using `=`. There is no `<-`

--

```{python}
# Python
a = 3
print(a + 2)
```

---

# <br> The Basics

<br>

## Basic types in R

--

```{r}
# R
# Note: factor and raw not included
logical <- TRUE
numeric <- 2.72
integer <- 42L
complex <- 3 + 2i
character <- "string"
cat(logical, numeric, integer, complex, character)
```

---

# <br> The Basics

<br>

## Basic types in Python

--

```{python}
# Python
logical = True
numeric = 2.72
integer = 42
complex = 3 + 2j # Note: j... like an engineer.
character = "string"
print(logical, numeric, integer, complex, character)
```

---

# <br> The Basics

<br>

## What about NA, NaN, Null, -Inf, and Inf?

--

Python has `None`

- `NA` and others need to be imported from packages. 
- It is roughly equivalent to `Null`.
- We'll get to these later.

--

"Use `is` when you want to check against an object's identity (e.g. checking to see if var is None). Use `==` when you want to check equality (e.g. Is var equal to 3?)." - [Stack Overflow User](https://stackoverflow.com/questions/14247373/python-none-comparison-should-i-use-is-or#14247383)

--

```{python}
#Python
a = None
print(a is None)
```

---

# <br> The Basics

<br>

## If Statements in R

--


```{r}
# R
confused <- FALSE
if (!confused) {
  cat("I'm getting it.")
} else {
  cat("Huh?")
}
```

---

# <br> The Basics

<br>

## if Statements in Python

--

```{python}
# Python
confused = False
if not confused:
    print("I'm getting it.")
else:
    print("Huh?")
```

---

# <br> The Basics

<br>

## `if-elif-else`

--

<br>

```{python}
# Python
confusion_level = 0 # assume confusion is 0 to 10 (very confused)
if 0 <= confusion_level < 3:
    print("I'm getting it.")
elif 3 <= confusion_level < 6:
    print("I'm kind of getting it.")
elif 6 < confusion_level <= 10:
    print("I'm not really getting it.")
else:
    print("Where am I?")
```

---

# <br> The Basics

<br>

## `for` Loops

```{R}
# R
for (i in c(1, 2, 3)) {
  print(i)
}
```

```{python}
# Python
for i in (1, 2, 3):
    print(i)
```

---

# <br> The Basics

<br>

## Another `for` Loop

```{R}
# R
for (i in seq_len(3)) {
  print(i)
}
```

```{python}
# Python
for i in range(3):
    print(i+1)
```

---

# <br> The Basics

<br>

## `while` Loops in R

```{R}
# R
i <- 1
while (i < 4) {
  print(i)
  i <- i + 1
}
```

---

# <br> The Basics

<br>

## `while` Loops in Python

```{python}
# Python
i = 1
while i < 4:
    print(i)
    i += 1
```

---

# <br> The Basics

<br>

## Functions

```{r}
# R
add_one <- function(v) {
  v + 1
}

cat(add_one(3))
```

```{python}
# Python
def add_one(v):
    return v + 1
    
print(add_one(3))
```

---

# <br> The Basics

<br>

## Don't forget to return values you want

```{python}
# Python
def add_one(v):
    ret = v + 1
    print(ret)
    ret
    
print(add_one(3))
```

---

# <br> The Basics

<br>

## Documenting your function

```{python}
# Python

def add_one(v):
    '''Add one to a vector of numeric values.'''
    ret = v + 1
    print(ret)
    ret

help(add_one)
```

---

# <br> The Basics

<br>

## R Lists

```{r}
# R
v <- list(1, 2)
print(v)
```

---

# <br> The Basics

<br>

## R Lists

```{r}
# R
v <- c(v, list(3))
print(v)
```

---

# <br> The Basics

<br>

## R Lists

```{r}
# R
print(v[-1])
```

---

# <br> The Basics

<br>

## Python Lists

```{python}
# Python
v = [1, 2]
print(v)
```

```{python}
# Python
v = [1, 2]
print(v + [3])
```

---

# <br> The Basics

<br>

## Python Lists Continued 

```{python}
# Python
v = [1, 2, 3]
print(v[-1])
```

---

# <br> The Basics

<br>

## List Indexing

<br>

R indexes start at 1. Python starts at 0.

```{r}
# R
v <- c('a', 'b', 'c')[3]
cat(v)
```

```{python}
# Python
v = ['a', 'b', 'c']
print(v[2])
```

---

# <br> The Basics

<br>

R indexes from the first argument _through_ the second. Python indexes from the first _up to_ the last (non-inclusive).

```{r}
# R
v <- c('a', 'b', 'c')[1:2]
cat(v)
```

```{python}
# Python
v = ['a', 'b', 'c']
print(v[0:2])
print(v[:2])
```

---

# <br> The Basics

<br>

## Lists containing values with different types

```{python}
# Python
list_vals = [1, '2', None]

def add_one(list_vals):
    for i in range(len(list_vals)):
        if list_vals[i] is None:
            print("Value at index " + str(i) + " is None. Not adding 1.")
        elif isinstance(list_vals[i], str):
            list_vals[i] = str(int(list_vals[i]) + 1)
        elif isinstance(list_vals[i], int):
            list_vals[i] += 1
        else:
            print("Unsupported type at ", i, ". Not doing anything")
    return list_vals

```

---

# <br> The Basics

<br>

## Lists containing values with different types (cont'd) 

```{python}
# Python
print(list_vals)
print(add_one(list_vals))
print(list_vals)
```

---

# <br> The Basics

<br>

## Values are passed _by reference_ 

```{python}
# Python
list_vals = [1, '2', None]

def add_one(list_vals):
    ret_vals = list_vals.copy()
    for i in range(len(ret_vals)):
        if ret_vals[i] is None:
            print("Value at index " + str(i) + " is None. Not adding 1.")
        elif isinstance(ret_vals[i], str):
            ret_vals[i] = str(int(ret_vals[i]) + 1)
        elif isinstance(ret_vals[i], int):
            ret_vals[i] += 1
        else:
            print("Unsupported type at ", i, ". Not doing anything")
    return ret_vals
```

---

# <br> The Basics

<br>

## Values are passed _by reference_ (cont'd) 

```{python}
# Python
lv_one_added = add_one(list_vals)
print(lv_one_added)
print(list_vals)
```

---

# <br> The Basics

<br>

## `=` Doesn't copy lists

```{python}
# Python
a = [3]
b = a
b[0] = 4
print(a)
```

---

# <br> The Basics

<br>

## It does copy singletons

```{python}
# Python
a = 3
b = a
b = 4
print(a)
```

---

# <br> The Basics

## Lists are not like R vectors 

```{r}
# R
a <- 1:3
print(a + 1)
```

```{python error = TRUE}
# Python
a = list(range(1, 4))
print(a)
a + 1
```

---

# <br> The Basics

<br>

## List Comprehensions

```{r}
# R
l <- as.list(1:3)
print(lapply(l, function(x) x + 1))
```

---

# <br> The Basics

<br>

## List Comprehensions

```{python}
# Python
l = list(range(1, 4))
print( [x + 1 for x in l] )
```

---

# <br> The Basics

<br>

## A Note on Tuples 

In Python a tuple is an _immutable_ list. 

```{python error = TRUE}
t1 = 1, 2, 3   # Create a tuple like this...
t2 = (1, 2, 3) # or like this

t1[1] = 2 # Throws an error

first_equal, second_equal, third_equal = [ t1[i] == t2[i] for i in range(len(t1)) ]
print(first_equal, second_equal, third_equal)
```

---

# <br> The Basics

<br>

## Dictionaries in Python are named lists in R) 

```{r}
# R
l <- list(a = c(1:3), b = c("c", "d", "e", "f"))
l$b[2:4]
```

```{python}
# Python
l = {'a' : list(range(1, 4)),
     'b' : ["c", "d", "e", "f"]
     }
     
print(l['b'][1:4])
```

---

# <br> The Basics

<br>

## Dictionaries are composed of keys and values

```{python}
# Python
l = {'a' : list(range(1, 4)),
     'b' : ["c", "d", "e", "f"]
    }

print(l.keys())
print(l.values())
```

---

```{css, echo = FALSE}
.huge {
  font-size: 200%;
}
```

<br>
<br>
<br>
<br>
.center[
.huge[
Great job. You made it to the end of part 1.
]
]
