---
title: 'Part 2: Digging Deeper into Python Programming Constructs'
author: "Michael Kane<br><span style = 'font-size: 80%;'>School of Public Health, Biostatistics Department<br>Yale University<br><br>"
date: '`r icon::fa("github")` [kaneplusplus](https://github.com/kaneplusplus)<br> `r icon::fa("twitter")` [kaneplusplus](https://twitter.com/kaneplusplus)'
output:
  xaringan::moon_reader:
    css: [duke-blue, metropolis, tamu-fonts]
    titleSlideClass: [left, middle]
---

```{r echo = FALSE}
library(reticulate)
xaringanExtra::use_panelset()
use_condaenv("python-for-r-developers", required = TRUE)
```

# <br> Highlights from Part 1


<br>

## 1. Python has a notion of environments which enacapsulate the interpreter 
## and a set of package.

--
<br>

## 2. Python syntax is distinct from, but not unrelated to R

--
<br>

## 3. Most of the time we've call functions with `function(object)` but 
## sometimes it's been `object.function()` (as with copy). We'll talk more 
## about this later.

--
<br>

## 4. Zero indexing.

---

# <br> Topics for Part 2

<br>

## 1. Packages

--
<br>

## 2. Numeric computing with Numpy.

--
<br>

## 3. Using Objects

---

# <br> Python Packages

--
<br>

## Python is "batteries not included."
--

<br>

## R includes _a lot_ of computing facilities with the core language.

--

<br> 

## 1. Plotting

--

## 2. Vectors, matrices, and arrays

--

## 3. Optimized and vectorized linear algebra routines

--

## 4. Suite of statistical functions and models

--

## 5. `data.frame`s
--

<br>

## None of these are included with Python.

---

# <br> Python Packages

--

<br>

## Python Virtual Environments and Package Management

--

<br>

## This class uses conda to create an environment and add packages. 

<br>

--

## [Anaconda](https://www.anaconda.com/) creates environments and adds packages from [Anaconda Cloud](https://anaconda.org/anaconda/repo). 
## It is an environment and package manager, not just for Python. 
<br>
--

## There is an R installation.
<br>
--

## I have heard only bad things about it.

---

# <br> Python Packages


<br>

## What's with the emphasis on environments? 

<br>

--

Development culture is different than R's

- R community (CRAN maintainers) place a higher value on package user's time by enforcing _downstream dependencies_. This is the reason R's `install.packages()` function "just works". It does a better job of hiding analysts from package development.

--

- Python community places higher value on developer time. It is often up to the user to sort out compatibility problems between packages. This became a big enough issue that companies, like [Continuum Analytics](http://www.continuumanalytics.com/), began creating pre-packaged virtual environments.

--

Result is that R users tend to use environments (with `packrat`, `switchr`, `renv`) more for reproducibility, Python users tend to use them more for package compatibility.

---

# <br> Numerical Computing 

<br>

## Python doesn't have a built-in notion of vectorized operations.

--

```{r echo = FALSE}
library(reticulate)
use_condaenv("r-reticulate")
```


```{python error = TRUE}
# Python
list(range(10)) + 1
```

--

We can perform this with list comprehensions. 

--

```{python error = TRUE}
# Python
print( [x + 1 for x in list(range(10))] )
```

--

Or, we could create a method for adding values to lists, but it's already been done.

---

# <br> Numerical Computing

<br>

## Importing the `numpy` Package

--

```{python}
# Python
import numpy
print( numpy.arange(10) + 1 )
```

--

```{python}
# Python
from numpy import *
print( arange(10) + 1 )
```

--

```{python}
# Python
import numpy as np
print( np.arange(10) + 1 )
```

---

# <br> Numerical Computing

<br>

## Numpy doesn't distinguish vectors, matrices, and arrays

```{python}
# Python

# A vector
vec = np.array(list(range(12)))
print(vec)

# A matrix
mat = np.array( [ [1, 2, 3], [4, 5, 6], [7, 8, 9], [10, 11, 12] ] )
print(mat)
```

---

# <br> Numerical Computing

<br>


## Numpy doesn't distinguish vectors, matrices, and arrays (cont'd) 

```{python}
# Python

tensor3 = np.array( [ [[1, 2], [3, 4]], [[5, 6], [7,8]] ])
print(tensor3)
```

---

# <br> Numerical Computing

<br>

## What is the dimension of a numpy array?

<br>

```{python}
# Python
print(vec.shape)
print(mat.shape)
print(tensor3.shape)
```

---

# <br> Numerical Computing

<br>

## What kind of values are stored?

```{python}
# Python

tensor3.dtype

double(tensor3).dtype

np.array([str(x) for x in mat.flatten().tolist()], dtype = str).reshape(4, 3)
```

---

# <br> Numerical Computing

<br>

## What other information is stored in a numpy array? 

```{python} 
# Python

vec_slots = dir(vec)
len(vec_slots)
print( [vec_slots[-i] for i in range(1, 51)] )
```

---

# <br> Numerical Computing

<br>

## What else is in numpy? 

```{python}
# Python

np_objects = dir(np)
len(np_objects)
print( [np_objects[i] for i in range(70)] )
```

---

# <br> Numerical Computing

<br>

## Array Indexing

```{r}
# R

mat <- t(matrix(seq_len(12), ncol = 4))
mat
mat[1:2, 1:3]
```

---

# <br> Numerical Computing

<br>

## Array Indexing (cont'd)

```{python}
# Python

mat[:2, :3]

mat[1:3, :3]

mat[ [0, 3, 2], :]
```

---

# <br> Numerical Computing

<br>

## Boolean Indexing

```{r}
# R

mat > 2
mat[mat > 2]
```

---


# <br> Numerical Computing

<br>

## Boolean Indexing (cont'd) 

```{python}
# Python

mat > 2
mat[mat > 2]
```

---

# <br> Numerical Computing

<br>


## Fitting Ordinary Least Squares

Recall the formula for fitting the ordinary least squares model:

$$
\begin{align}
\widehat{\beta} &= (X^T \ X)^{-1} \ X^T \ Y. \\
\end{align}
$$
Letting $X = QR$ where $Q^TQ = I$ and $R$ is upper right triangular we can rewrite as:

$$
\begin{align}
\widehat{\beta} &= ( (QR) ^T\  QR) ^{-1} \ (Q R)^{T} \ Y \\
&= (QR)^{-1} \ ((QR)^T )^{-1} \ (Q R)^{T} \ Y \\
&= (QR)^{-1} \ Y \\
&= R^{-1} Q^T Y
\end{align}
$$
to create a _numerically stable_, if not limited, implementation of OLS.

---

# <br> Numerical Computing

<br>

## Our implementation

```{python}
# Python

import seaborn as sns # for iris

def ols(Y, X):
    q, r = np.linalg.qr(X)
    return(np.linalg.inv( r ).dot( q.T ).dot( Y ))

iris = sns.load_dataset("iris")
iris_mat = iris[["sepal_width", "petal_length", "petal_width"]].values

print(ols(iris['sepal_length'].values, iris_mat))
```

---

# <br> Numerical Computing

<br>


## Our implementation (cont'd)

```{r}
fit <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width - 1, 
          data = iris)
fit$coefficients
```

---

# <br> Numerical Computing

<br>

## How would you debug this?

<br>

## Python doesn't have error recover the way R does 
## (`options(error = recover)`)

<br>

## It does let you set breakpoints with the pdb package though.

<br>

## [pdb-debug.py](pdb-debug.py)

---

# <br> Using Objects to Visualize Data

<br>

## Object Oriented Programming

You've already been making use of Python's object oriented functionality.

  - `list_vals.copy()`
  - `np.linalg.inv( r ).dot( q.T )`
  
In each case you were accessing data or calling a function called a (method) associated with an object using the `.` operator.

Packages are themselves object.

### An _object_ contains data (called attributes or fields) and methods (functions).

An object _has_ attributes and can do things with methods.

An object is an _instance_ of a class.

- `np` is an instance of type `Module`.
- `list_vals` is an instance of type `list`.

---

# <br> Using Objects to Visualize Data

## A Primitive Vector Object in R

```{r}
# R

vec_vals <- list(
  vals = 1:10,
  add_one = function(vec_vals) {
    vec_vals$vals <- vec_vals$vals + 1
    vec_vals
  }
)

print(vec_vals)
```

---

# <br> Using Objects to Visualize Data

<br>

## A Primitive Vector Object in R (cont'd)

```{r}
# R
print(vec_vals$add_one(vec_vals))

```

Note two differences:

1. Python uses `.` instead of our `$`. 
2. The calling object is invisibly passes as the first argument.

---

# <br> Using Objects to Visualize Data

<br>

## Plotting with Objects

```{r fig.height=3.5, fig.width=6}
# R
library(ggplot2)

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point() +
  theme_minimal()
```

---

# <br> Using Objects to Visualize Data

<br>

## Plotting with Objects

```{python, echo = FALSE}
import warnings
warnings.filterwarnings("ignore")

import plotnine
from plotnine import *
```

```{python eval = FALSE, warning = FALSE}
# Python

import plotnine
from plotnine import *

plotnine.options.figure_size = (5, 3)

p = ggplot(iris, aes(x = "sepal_length", y = "sepal_width", color = "species")) +\
  geom_point() + theme(subplots_adjust={'right': .7})
p.draw()
```

---

# <br> Using Objects to Visualize Data

<br>

## What's with `+\` ?

<br>

### Python expects something to the left and right of `+\`.

<br>

### It can't find the second argument on the second line.

<br>

### So we need to tell the interpreter to look at the next line with `\`.

<br>

### Alternative is to wrap the entire statement in `()`


```{python eval = FALSE}
(ggplot(iris, aes(x = "sepal_length", y = "sepal_width", color = "species")) +
    geom_point() + theme(subplots_adjust={'right': .7}))
```

---

# <br> Using Objects to Visualize Data

<br>

## Where are the objects?

The `ggplot` function creates an object of type `ggplot`.

```{python warning = FALSE}
# Python

p = ggplot(iris, aes(x = "sepal_length", y = "sepal_width", color = "species"))
type(p)
```

---

# <br> Using Objects to Visualize Data

<br>

## Where are the methods?

`+` is an _infix operator_ 

- a function where the first argument is on the left of the operator and the second is on the right
- it is implemented via the `__add__` method.

```{python warning = FALSE}
# Python

dir(p)
```

---

# <br> Using Objects to Visualize Data

<br>

## We can call `__add__` with `.`

```{python, warning = FALSE}
# Python

p = ggplot(iris, aes(x = "sepal_length", y = "sepal_width", color = "species"))

geom = geom_point()
type(geom)
theme_change = theme(subplots_adjust={'right': .7})

# Note that __add__ creates copies.
p = p.__add__(geom)
p = p.__add__(theme_change)
```

---

# <br> Using Objects to Visualize Data

<br>

## We can call `__add__` with `.` (cont'd)

<br>

```{python eval = FALSE, warning = FALSE}
# Python

p.draw()
```

# <br> Summing the section up

<br>

## Python includes less functionality and relies more on packages than R.

<br>

## Everything in Python is an object

<br>

## Object have functions and data that are accessed with `.`

<br>

## Those functions and data can be found with `dir()`

<br>

## Their documentation can be found with `help()`

---

```{css, echo = FALSE}
.huge {
  font-size: 200%;
}
```

<br>
<br>
<br>
<br>
.center[
.huge[
You made it to the end of part 2.
]
]
