---
title: 'Part 4: Working with Data'
author: "Michael Kane<br><span style = 'font-size: 80%;'>School of Public Health, Biostatistics Department<br>Yale University<br><br>"
date: '`r icon::fa("github")` [kaneplusplus](https://github.com/kaneplusplus)<br> `r icon::fa("twitter")` [kaneplusplus](https://twitter.com/kaneplusplus)'
output:
  xaringan::moon_reader:
    css: [duke-blue, metropolis, tamu-fonts]
    titleSlideClass: [left, middle]
---

```{r echo = FALSE}
library(reticulate)
xaringanExtra::use_panelset()
use_condaenv("python-for-r-developers", required = TRUE)
```

# <br> Overview of Part 4

--
<br>

## Highlights from Part 3

--
<br>

### Generators

--
<br>

### Classes and Objects 

---

# <br> Overview of Part 4 

<br>

### Importing Data with Pandas

--
<br>

### Data Cleaning

--
<br>

### Data Exploration

--
<br>

### Pricing Wine

--
<br>

### R's R6 Object System

--
<br>

### Calling Python from R

---

# <br> Importing Data with Pandas 

<br>

```{python}
# Python 
import numpy as np
import pandas as pd
import plotnine
from plotnine import *

plotnine.options.figure_size = (4, 2)

wine = pd.read_csv("winemag-data-130k-v2.csv")

wine = wine.drop('Unnamed: 0', axis = 1)

# How many rows?
wine.shape[0]
```

---

# <br> Importing Data with Pandas

<br>

## Accessing columns

--

```{python}
# Python

# What are the column names?
wine.keys()
```

---

# <br> Data Cleaning 

<br>

## Dealing with Missing Data

--

```{python}
# How many missing countries?
wine[ ~ wine['country'].isnull() ].shape[0]

# Drop them
wine = wine[ ~ wine['country'].isnull() ]
wine.shape
```

---

# <br> Data Cleaning

<br>

## How many wineries and countries?

--

```{python}
# How many unique wineries?
len(wine['winery'].unique())

# How many countries?
len(wine['country'].unique())
```

---

# <br> Data Exploration 

<br>

## Visualizing Country Counts

--

```{python, eval = FALSE}

# Make country categorical, ordered by count.
country_list = wine['country'].value_counts().index.tolist()
country_cat = pd.Categorical(wine['country'], categories = country_list)
wine = wine.assign(country_cat = country_cat)

# Where are the wines from?
p = (ggplot(wine, aes(x = "country_cat")) +
      theme(axis_text_x=element_text(rotation=270, hjust=1)) +
      geom_histogram())
p.draw()
```

---

# <br> Data Exploration 

<br>

## Where do the wines come from?

--

```{python, echo = TRUE}

# Make country categorical, ordered by count.
country_list = wine['country'].value_counts().index.tolist()
country_cat = pd.Categorical(wine['country'], categories = country_list)
wine = wine.assign(country_cat = country_cat)

# Where are the wines from?
p = (ggplot(wine, aes(x = "country_cat")) +
      theme(axis_text_x=element_text(rotation=270, hjust=1)) +
      geom_histogram() +
      theme(subplots_adjust={'bottom': .3}))

# p.draw()
```

---

# <br> Data Exploration 

<br>

## How much do they cost?

--

```{python}
p = (ggplot(wine, aes(x = "price")) +
        theme(axis_text_x=element_text(rotation=270, hjust=1)) +
        geom_histogram() +
        theme(subplots_adjust={'bottom': .3}))

# p.draw()
```

---

# <br> Data Exploration

<br>

## Switching to the log scale 

--

```{python, echo = TRUE, warning = FALSE, message = FALSE}
import warnings
warnings.filterwarnings("ignore")

# Let's log. Get rid of NaN's first.
wine = wine[ ~ wine['price'].isna() ]
wine = wine.assign(log_price = np.log(wine['price']))

p = (ggplot(wine, aes(x = "price")) +
    scale_x_continuous(trans='log2') +
    theme(axis_text_x=element_text(rotation=270, hjust=1)) +
    geom_histogram() +
    xlab("Price") +
    ylab("Count") +
    theme(subplots_adjust={'left': .1, 'bottom': 0.2}))
    
# p.draw()
```

---

# <br> Data Exploration 

<br>

## Wine Varieties

--

```{python}
# How many types are represented?
variety_count = pd.value_counts(wine.variety.values)

# How many varieties have been rated at least 1000 times?
np.sum(variety_count > 1000)
variety_count = variety_count[variety_count > 1000]

wine = wine[wine.variety.isin(variety_count.keys())]
wine.shape
```

---

# <br> Data Exploration 

## Wine Rating

--

```{python}
# What is the average number points per variety?
(wine.points.groupby(wine.variety).describe().sort_values("mean",
  ascending = False))[:10]
```

---

# <br> Data Exploration 

<br>

## Who are the wine tasters?

```{python}

# How many tasters?
taster_counts = pd.value_counts(wine['taster_name'].values)
taster_counts[:10]
```

---

# <br> Data Exploration

<br>
## Subsetting / Filtering

--

```{python, eval = TRUE}
# Let's get the individuals with at least 1000 tastings.
keep_tasters = taster_counts[taster_counts > 1000]

wine = wine[wine.taster_name.isin(keep_tasters.keys())]

(wine.points.groupby(wine.taster_name).describe().sort_values("mean",
  ascending = False))[:5]
```

---

# <br> Pricing Wine

<br>

## A Linear Regression

--

```{python, echo = TRUE}
# What is the assocation between price and points?
import statsmodels.formula.api as sm

fit = sm.ols(formula = "points ~ price", data = wine).fit()

ols_summary = fit.summary() # Try this yourself to see the output
```

---

# <br> Pricing Wine

<br>

## Another Linear Regression

--

```{python, eval = TRUE}
np.sort(wine.variety.unique())

# Is there an interaction with variety?
fit = sm.ols(formula = "points ~ log_price + variety + price:variety", \
  data = wine).fit()
ols_summary = fit.summary() # Ommited for slide real estate.
```

---

# <br> Pricing Wine

<br>

--

## Which wines are the best buy?

```{python}
# Which wine is the most overpriced?
wine = wine.assign(resid = fit.resid)
wine = wine.assign(fitted = fit.fittedvalues)
under_priced = wine[wine.resid == np.max(wine.resid)]
under_priced.keys()
```

---

# <br> Pricing Wine

<br>

## Which wines are the best buy (cont'd)?

```{python}
under_priced.values
```

---

# <br> Pricing Wine

<br>

## Which ones are the worst?

```{python}
over_priced = wine[wine.resid == np.min(wine.resid)]
over_priced.values
```

---

# <br> R's R6 Object System

<br>

--

## Python's object are standard in object oriented programming

<br>

--

## R has a similar system called R6

<br>

--

## Methods and attributes are associated with an environment

<br>

--

## Instead of `.` , we use `$`.

---


# <br> R's R6 Object System

--

```{r}
library(R6)
AddOneToList <- R6Class(
  "AddOneToList",
  public = list(
      initialize = function(lst) {
        private$lst <- lst
        self$greet()
      },
      add_one = function() {
        stop("Method is abstract")
      },
      get_lst = function() {
        stop("Method is abstract")
      },
      greet = function() {
        cat(paste0("Abstract add one to list created", ".\n"))
      }
    ),
  private = list(
    lst = NULL
  )
)
```

---

# <br> R's R6 Object System

<br>

## Let's try to instantiate an object

<br>

```{r error = TRUE}
lst_adder <- AddOneToList$new(as.list(1:5))
lst_adder$add_one()
```

---

# <br> R's R6 Object System

```{r}
AddOneToNumericList <- R6Class("AddOneToNumericList",
  inherit = AddOneToList,
  public = list(
    initialize = function(lst) {
      #browser()
      if (!all(is.numeric(unlist(lst)))) {
        stop("All values should be numeric.")
      }
      super$initialize(lst)
    },
    add_one = function() {
      private$lst <- lapply(private$lst, function(x) x + 1)
    },
    get_lst = function() {
      private$lst
    },
    greet = function() {
      cat(paste0("Concrete add one to list created", ".\n"))
    }
  )
)
```

---

# <br> R's R6 Object System

<br>

## Let's instantiate a proper concrete object

```{r error = TRUE}
lst_adder <- AddOneToNumericList$new(as.list(letters[1:5]))

lst_adder <- AddOneToNumericList$new(as.list(1:5))

lst_adder$add_one()

unlist(lst_adder$get_lst())
```
---

# <br> Calling Python from R

<br>

--

## R mirrors Pythons object through the R6 class system

<br>

--

## We've created Python code that does something useful

<br>

--

- We'd like to be able to reuse the code and objects created.

--

<br>

- We'd like to use R to augment our analysis.


---

# <br> Calling Python from R 

<br>

## Sourcing a Python file
```{r}
# R
source_python("part-4.py", convert = FALSE)

ls()[1:10]
```

---

# <br> Calling Python from R

<br>

## Get the column names of the wine data set into R

```{r message = FALSE, warning = FALSE}
library(dplyr)

col_keys <- wine$keys()$to_list() %>% 
  py_to_r()

col_keys
```

---

# <br> Calling Python from R

## Now read the Pandas data frame into R as a `tibble`


```{r}
# R
library(tibble)
library(DT)

handle_nan_and_unlist <- function(x) {
  x <- unlist(x)
  x[is.nan(x)] <- NA
  x
}

wine_df <- wine[as.list(setdiff(col_keys, "country_cat"))] %>%
  py_to_r() %>%
  mutate_if(is.list, handle_nan_and_unlist) %>%
  as_tibble()
```

---

# <br> Calling Python from R

## Now read the Pandas data frame into R as a `tibble`


```{r}
wine_df %>% 
  head(n = 100) %>% 
  datatable()
```

---

# <br> Calling Python from R

<br>

## What about our numerical routines?

```{r}
import("numpy", as = "np")

ols_code <- '
def ols(Y, X):
  q, r = np.linalg.qr(X)
  return(np.linalg.inv( r ).dot( q.T ).dot( Y ))
'

writeLines(ols_code, "ols_code.py")
source_python("ols_code.py", convert = FALSE)
```

---

# <br> Calling Python from R

## What about our numerical routines?

```{r}
form <- points ~ log_price + variety + price:variety
wine_mf <- model.frame(form, wine_df) 
X <- model.matrix(form, wine_mf)
Y <- matrix(wine_mf$points, ncol = 1)

fit <- ols(X, Y)$flatten() %>%  py_to_r()
names(fit) <- colnames(X)
fit[1:10]
```

---

# <br> Wrapping up

<br>

## Python has modeled most of it's data science libraries after R

<br>

## Python's programming constructs although distinct provide a new twist 
## on familar method of programming

<br>

## If you can understand the syntax you can figure out how to use the libraries

<br>

## If you want, you can do this from R

---

```{css, echo = FALSE}
.huge {
  font-size: 200%;
}
```

<br>
<br>
<br>
<br>
.center[
.huge[
You made it to the end of the class. Thanks very much!
]
]

