---
title: 'Part 2: Digging Deeper into Python Programming Constructs'
author: "Michael Kane, Yale University"
output: 
  ioslides_presentation:
    theme: flatly
    smaller: true
---

```{r echo = FALSE}
library(reticulate)
use_condaenv("r-reticulate")
```

# Overview of Part 2 {.smaller}

## Highlights from Part 1

### 1. Python has a notion of environments which enacapsulate the interpreter and a set of package

### 2. Python syntax is distinct from, but not unrelated to R

### 3. Most of the time we've call functions with `function(object)` but sometimes it's been `object.function()` (as with copy). We'll talk more about this later.

### 4. Zero indexing.

## Topics for this part

### 1. Packages

### 2. Numeric computing with Numpy.

### 3. Using Objects

### 3. Generators

# Packages

## Python is "batteries not included"

R includes _a lot_ of computing facilities with the core language.

1. Plotting
2. Vectors, matrices, and arrays
3. Optimized and vectorized linear algebra routines
4. Suite of statistical functions and models
5. `data.frame`s

None of these are included with Python.

## Python Virtual Environments and Package Management

This class uses virtualenv to create an environment and pip3 within the environment to add packages. Packages are install from [PyPI](https://pypi.org/).

[Anaconda](https://www.anaconda.com/) is another option for both creating environments and adding packages from [Anaconda Cloud](https://anaconda.org/anaconda/repo). It is an environment and package manager, not just for Python. There is an R installation, I have heard only bad things about this build.

## What's with the emphasis on environments? {.smaller}

Development culture is different than R's

- R community (CRAN maintainers) place a higher value on package user's time by enforcing _downstream dependencies_. This is the reason R's `install.packages()` function "just works". It does a better job of hiding analysts from package development.
- Python community places higher value on developer time. It is often up to the user to sort out compatibility problems between packages. This became a big enough issue that companies, like [Continuum Analytics](http://www.continuumanalytics.com/), began creating pre-packaged virtual environments.

Result is that R users use environments (with `packrat`, `switchr`, `renv`) for reproducibility, Python users use them for reproducibility and package compatibility.

## A second look at setup-p4rp.sh {.smaller}

```{bash, eval = FALSE}
# Make sure python 3 and pip3 are installed on you machine and accessible from the
# command line.

# Install virtualenv.
pip3 install virtualenv

# Create a "python for R programmer's" virtual environment.
virtualenv -p python3 p4rp

# Activate the virtual environment. Note: this works for Mac and Linux.
source p4rp/bin/activate

# The following *may* work on Windows.
# p4rp\bin\activate.bat

# Install the needed packages.
pip3 install numpy pandas seaborn plotnine
```

## Numeric Computing {.smaller}

Python doesn't have a built-in notion of vectorized operations.

```{r echo = FALSE}
library(reticulate)
use_virtualenv("./p4rp")
```

```{python error = TRUE}
# Python
list(range(10)) + 1
```

We can perform this with list comprehensions. 

```{python error = TRUE}
# Python
print( [x + 1 for x in list(range(10))] )
```

Or, we could create a method for adding values to lists, but it's already been done.

## Importing the `numpy` Package {.smaller}

```{python}
# Python
import numpy
print( numpy.arange(10) + 1 )
```

```{python}
# Python
from numpy import *
print( arange(10) + 1 )
```

```{python}
# Python
import numpy as np
print( np.arange(10) + 1 )
```

## It's all arrays {.smaller}

Numpy doesn't distinguish vectors, matrices, and arrays.

```{python}
# Python

# A vector
vec = np.array(list(range(12)))
print(vec)

# A matrix
mat = np.array( [ [1, 2, 3], [4, 5, 6], [7, 8, 9], [10, 11, 12] ] )
print(mat)
```

## It's all arrays (cont'd) {.smaller}

```{python}
# Python

tensor3 = np.array( [ [[1, 2], [3, 4]], [[5, 6], [7,8]] ])
print(tensor3)
```

## What is the dimension of a numpy array?

```{python}
# Python
print(vec.shape)
print(mat.shape)
print(tensor3)
```

## How are values stored?

Check the `dtype`.

```{python}
# Python

tensor3.dtype

double(tensor3).dtype

np.array([str(x) for x in mat.flatten().tolist()], dtype = str).reshape(4, 3)
```

## What other information is stored in a numpy array? {.smaller}

```{python} 
# Python

vec_slots = dir(vec)
len(vec_slots)
print( [vec_slots[-i] for i in range(1, 51)] )
```

## What else is in numpy? {.smaller}

```{python}
# Python

np_objects = dir(np)
len(np_objects)
print( [np_objects[i] for i in range(70)] )
```

## Array Indexing {.smaller}

```{r}
# R

mat <- t(matrix(seq_len(12), ncol = 4))
mat
mat[1:2, 1:3]
```

## Array Indexing (cont'd) {.smaller}

```{python}
# Python

mat[:2, :3]

mat[1:3, :3]

mat[ [0, 3, 2], :]
```

## Boolean Indexing {.smaller}

```{r}
# R

mat > 2
mat[mat > 2]
```

## Boolean Indexing (cont'd) {.smaller}

```{python}
# Python

mat > 2
mat[mat > 2]
```

## Fitting Ordinary Least Squares

Recall the formula for fitting the ordinary least squares model:

$$
\begin{align}
\widehat{\beta} &= (X^T \ X)^{-1} \ X^T \ Y. \\
\end{align}
$$
Letting $X = QR$ where $Q^TQ = I$ and $R$ is upper right triangular we can rewrite as:

$$
\begin{align}
\widehat{\beta} &= ( (QR) ^T\  QR) ^{-1} \ (Q R)^{T} \ Y \\
&= (QR)^{-1} \ ((QR)^T )^{-1} \ (Q R)^{T} \ Y \\
&= (QR)^{-1} \ Y \\
&= R^{-1} Q^T Y
\end{align}
$$
to create a _numerically stable_, if not limited, implementation of OLS.

## Our implementation

```{python}
# Python

import seaborn as sns # for iris

def ols(Y, X):
    q, r = np.linalg.qr(X)
    return(np.linalg.inv( r ).dot( q.T ).dot( Y ))

iris = sns.load_dataset("iris")
iris_mat = iris[["sepal_width", "petal_length", "petal_width"]].values

print(ols(iris['sepal_length'].values, iris_mat))
```

## Our implementation (cont'd)

```{r}
fit <- lm(Sepal.Length ~ Sepal.Width + Petal.Length + Petal.Width - 1, 
          data = iris)
fit$coefficients
```

# Using Objects to Visualize Data

## Object Oriented Programming

You've already been making use of Python's object oriented functionality.

  - `list_vals.copy()`
  - `np.linalg.inv( r ).dot( q.T )`
  
In each case you were accessing data or calling a function called a (method) associated with an object using the `.` operator.

Packages are themselves object.

### An _object_ contains data (called attributes or fields) and methods (functions).

An object _has_ attributes and can do things with methods.

An object is an _instance_ of a class.

- `np` is an instance of type `Module`.
- `list_vals` is an instance of type `list`.

## A Primitive Vector Object in R

```{r}
# R

vec_vals <- list(
  vals = 1:10,
  add_one = function(vec_vals) {
    vec_vals$vals <- vec_vals$vals + 1
    vec_vals
  }
)

print(vec_vals)
```

## A Primitive Vector Object in R (cont'd)

```{r}
# R
print(vec_vals$add_one(vec_vals))

```

Note two differences:

1. Python uses `.` instead of our `$`. 
2. The calling object is invisibly passes as the first argument.

## Plotting with Objects

```{r fig.height=3.5, fig.width=6}
# R

library(ggplot2)

ggplot(iris, aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point()
```

## Plotting with Objects

```{python, echo = FALSE}
import warnings
warnings.filterwarnings("ignore")
```

```{python eval = FALSE, warning = FALSE}
# Python

import plotnine
from plotnine import *

plotnine.options.figure_size = (2, 1)

ggplot(iris, aes(x = "sepal_length", y = "sepal_width", color = "species")) +\
  geom_point() + theme(subplots_adjust={'right': .7})
```

## Plotting with Objects (cont'd)

```{python warning = FALSE, echo = FALSE}
# Python

import plotnine
from plotnine import *

plotnine.options.figure_size = (4, 2)

ggplot(iris, aes(x = "sepal_length", y = "sepal_width", color = "species")) +\
  geom_point() + theme(subplots_adjust={'right': .7})
```

## What's with `+\`

### Python expects something to the left and right of `+\`.

### It can't find the second argument on the second line.

### So we need to tell the interpreter to look at the next line with `\`.

### Alternative is to wrap the entire statement in `()`

```{python eval = FALSE}
(ggplot(iris, aes(x = "sepal_length", y = "sepal_width", color = "species")) +
    geom_point() + theme(subplots_adjust={'right': .7}))
```

## Where are the objects?

The `ggplot` function creates an object of type `ggplot`.

```{python warning = FALSE}
# Python

p = ggplot(iris, aes(x = "sepal_length", y = "sepal_width", color = "species"))
type(p)
```

## Where are the methods?

`+` is an _infix operator_ 

- a function where the first argument is on the left of the operator and the second is on the right
- it is implemented via the `__add__` method.

```{python warning = FALSE}
# Python

dir(p)
```

## We can call `__add__` with `.`

```{python, warning = FALSE}
# Python

p = ggplot(iris, aes(x = "sepal_length", y = "sepal_width", color = "species"))

geom = geom_point()
type(geom)
theme_change = theme(subplots_adjust={'right': .7})

# Note that __add__ creates copies.
p = p.__add__(geom)
p = p.__add__(theme_change)
```

## We can call `__add__` with `.` (cont'd)

```{python, warning = FALSE}
# Python

p.draw()
```
